---
title: "Wearables_analysis"
format: html
editor: visual
---

# 1. Resumen de la tarea empresarial

Análisis de tendencias de uso de productos Fitbit con el fin de responder a las necesidades de los usuarios y dirigir campañas de marketing de la empresa competidora [Bellabeat](https://bellabeat.com/).

Nos centramos en un sólo producto de Bellabeat:

- *Leaf*: Dispositivo de seguimiento clásico de bienestar de Bellabeat que se puede usar como pulsera, collar o clip. El dispositivo Leaf se conecta a la aplicación Bellabeat para hacer un seguimiento de la actividad física, el sueño y el estrés.

Este es el producto que mejor se adapta a la información que podemos obtener con los datos de Fitbit ya que tenemos los minutos dormidos y la actividad diaria. El estrés es difícil de operacionalizar con los datos que tenemos.

# 2. Descripción de todas las fuentes de datos utilizadas

[Datos de seguimiento de actividad física de Fitbit](https://www.kaggle.com/datasets/arashnic/fitbit). Este conjunto de datos de Kaggle contiene el seguimiento de la actividad física personal en treinta usuarios de Fitbit. Treinta usuarios elegibles de Fitbit prestaron su consentimiento para el envío de datos personales de seguimiento que incluyen rendimiento de la actividad física en minutos, ritmo cardíaco y monitoreo del sueño. Incluye información sobre la actividad diaria, pasos y ritmo cardíaco que se puede usar para explorar los hábitos de los usuarios. Esta información fue recogida a través de una encuesta en la plataforma Amazon Mechanical Turk entre el 12 de marzo de 2016 y 12 de mayo del mismo año. Es decir, 2 meses.

Recordar explicar que algunas definiciones de variables y valores sólo están  [aquí](https://dev.fitbit.com/build/reference/web-api) 

Recordar explicar lo de que no tenemos demographic data

Hay problemas con el sesgo o la credibilidad de estos datos?

● ¿Cómo estás abordando la autorización, la privacidad, la seguridad y la accesibilidad?

**Diccionario de variables**

LogId: ID of the sleep log to be deleted.

# 3. Documentación de todas las limpiezas y manipulaciones de datos

● ¿Cómo verificaste la integridad de los datos? ● ¿Existe algún problema con los datos?

## 3.1. Libraries

```{r libraries}
library(tidyverse)
library(DataExplorer)
library(lubridate)

options(scipen = 999)#to not show scientific notation
```

## 3.2. Data

Recordar explicar en el readme que saqué una carpeta hacia arriba los datos porque era redundante y que renombre a data la carpeta principal. Y que la carpeta principal está en gitignore.

```{r data}
dailyActivity_merged1<-read_csv(file = "./data/Fitabase Data 3.12.16-4.11.16/dailyActivity_merged.csv")
dailyActivity_merged2<-read_csv(file = "./data/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv")

minuteSleep_merged1<-read_csv(file = "./data/Fitabase Data 3.12.16-4.11.16/minuteSleep_merged.csv")
minuteSleep_merged2<-read_csv(file = "./data/Fitabase Data 4.12.16-5.12.16/minuteSleep_merged.csv")



head(dailyActivity_merged1)
head(dailyActivity_merged2)

head(minuteSleep_merged1)
head(minuteSleep_merged2)
```

## 3.3. Cleaning

- Joining dataset from different dates
- Selecting variables
- Converting dataset into long format (Explain more what that means later)
- Changing variables format
- Giving variables descriptive names
- Giving values descriptive and easy to read names

```{r joinActivity}
dailyActivity<- dailyActivity_merged1 |> full_join(dailyActivity_merged2) |> 
  select(Id, Date=ActivityDate, ends_with("Minutes")) |> 
  pivot_longer(
    cols = ends_with("Minutes"),
    names_to = "Activity_intensity",
    values_to = "Minutes"
  ) |> 
  mutate(
    Activity_intensity = str_remove(Activity_intensity, "Minutes"),
    Activity_intensity = str_remove(Activity_intensity, "Active"),
    Date= mdy(Date)
         )

dailyActivity
```

```{r joinSleep}
minuteSleep<- 
  minuteSleep_merged1 |> full_join(minuteSleep_merged2) |> 
  select(Id, Date=date,Sleep_state=value, LogId=logId) |> 
  mutate(
    # I don't want one user to seem twice the other. These are quali
    Id = as.character(Id),
    LogId = as.character(LogId),
    
    Date_time = mdy_hms(Date), # Handles "4/12/2016 2:47:30 AM"
    Date = as.Date(Date_time),      # Drops the time, keeps YYYY-MM-DD
      
    # Change 1, 2, 3 into readable categories
    Sleep_state = case_when(
      Sleep_state == 1 ~ "Asleep",
      Sleep_state == 2 ~ "Restless",
      Sleep_state == 3 ~ "Awake"
        )
    ) 

minuteSleep
```

```{r sleepWide}
sleepWide<-
  minuteSleep |> group_by(Id, Date, Sleep_state) |> 
  # 4. Count the rows (1 row = 1 minute)
  summarize(total_minutes = n(), .groups = "drop")|> 
  # 5. Pivot wider to create separate columns for Asleep, Restless, and Awake
  pivot_wider(
    names_from = Sleep_state,
    values_from = total_minutes,
    values_fill = 0 # Puts a 0 if the user had 0 minutes in a state that day
  ) 

sleepWide
```

```{r joinUsers}
#  CREO QUE ESTO NO AYUDA
# iba a hacer algo con lo de que hay más usuarios en activity q sleep pero da igual
# Users<-
  # dailyActivity |> full_join(minuteSleep)

```

# 4. Resumen de tu análisis

## Análisis exploratorio

### Actividad

```{r usersActivity}
dailyActivity |> distinct(Id) |> count(name="n_users")
```

```{r histActivity}
dailyActivity |> pull(Minutes) |> hist()
```

```{r}
dailyActivity |>group_by(Activity_intensity) |> summarise(sum(Minutes))
```

```{r}
summary_activity<-dailyActivity |>group_by(Activity_intensity) |> summarise(Minutes=sum(Minutes)) 

ggplot(summary_activity, aes(x=Activity_intensity, y=Minutes)) + 
  geom_bar(stat = "identity")
```


```{r}
dailyActivity |>group_by(Id,Activity_intensity) |> summarise(sum(Minutes))
```

```{r}
dailyActivity |>group_by(Id) |> summarise(sum(Minutes))
```

### Sueño

```{r usersSleep}
minuteSleep |> distinct(Id) |> count(name="n_users")
```

Menos que los 35 usuarios de activity

```{r stateDistribution}
minuteSleep |> count(Sleep_state)
```

```{r userSleep}
minuteSleep |> filter(Id==1503960366, Date=="2016-03-17") |> select(-Date)
```

```{r sleelSchedule}
sleep_schedule <-
  minuteSleep |> 
  # Group by the specific sleep session (logId) and the user
  group_by(Id, LogId) |> 
  
  # Find the first time (min) and last time (max) of that session
  summarize(
    bedtime = min(Date_time),
    wake_time = max(Date_time),
    .groups = "drop"
  ) |> 
  mutate(
    # Calculate total sleep time in hours
    duration_hours = as.numeric(difftime(wake_time, bedtime, units = "hours")),
    
    # Extract the day of the week the sleep session started
    # label = TRUE gives us names (Monday, Tuesday) instead of numbers
    day_of_week = wday(bedtime, label = TRUE, abbr = FALSE) 
  )|> 
  relocate(duration_hours, .before = bedtime)

sleep_schedule
```


```{r sleepTimes}
# See the most common bedtimes (in 24-hour format)
sleep_schedule |>   
  # Extract just the hour to make it easy to find trends
  mutate(
    bed_hour = hour(bedtime),
    wake_hour = hour(wake_time)
  ) |> 
  count(bed_hour, sort = TRUE)

# See the most common wake times
sleep_schedule |>   
  # Extract just the hour to make it easy to find trends
  mutate(
    bed_hour = hour(bedtime),
    wake_hour = hour(wake_time)
  ) |> 
  count(wake_hour, sort = TRUE)

# See the most common durations of sleep
sleep_schedule |> 
  count(hours_of_sleep=round(duration_hours,0), sort = TRUE)

# durations of sleep summary
summary(sleep_schedule$duration_hours)
```

```{r bxpltDuration}
ggplot(sleep_schedule, aes(x = "All Users", y = duration_hours)) +
  # outlier.shape = NA evita que los outliers se dibujen por duplicado
  geom_boxplot(fill = "#3498db", alpha = 0.7, color = "black", outlier.shape = NA) +
  
  # Añadimos los puntos con jitter. 
  # width = 0.2 controla cuánto se dispersan horizontalmente
  # alpha = 0.4 hace los puntos semitransparentes para ver dónde se acumulan
  geom_jitter(width = 0.2, alpha = 0.4, color = "darkblue") +
  
  # 8-hour benchmark for sleep
  geom_hline(yintercept = 8, linetype = "dashed", color = "#e74c3c", linewidth = 1) +
  
  labs(
    title = "Overall Distribution of Sleep Duration",
    subtitle = "Combined data for all users across all days",
    x = NULL, # Removes the x-axis title since "All Users" is self-explanatory
    y = "Hours of sleep"
  ) +
  theme_minimal()

```

Seems low but I am adding into the boxplot the naps and those should not last 8 hours. Also, how much can I trust the values if there are people sleeping 10 hour and more.


```{r hstDuration}
ggplot(sleep_schedule, aes(x = duration_hours)) +

  # binwidth = 0.5 is crucial; it sets each bar to represent a 30-minute window.
  geom_histogram(
    binwidth = 0.5,
    fill = "#3498db",  
    color = "black",
    alpha = 0.8 
  ) +
  
  # Reference line (8-hour benchmark)
  geom_vline(
    xintercept = 8,
    linetype = "dashed",
    color = "#e74c3c", 
    linewidth = 1
  ) +
  
  # 4. Enhance readability and formatting
  scale_x_continuous(
    breaks = seq(0, max(sleep_duration$duration_hours, na.rm = TRUE) + 1, by = 1)
    ) +
  labs(
    title = "Sleep Duration Distribution",
    subtitle = "Histogram showing how often specific sleep lengths occur (in 30-minute bins)",
    x = "Total Sleep Duration (Hours)",
    y = "Frequency of sleep duration"
  ) +
  theme_minimal() +
  # Make the title and axes text clear for presentation
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, face = "italic"),
    axis.title = element_text(size = 11)
  )
```
Most people are sleeping around 8 hours a night more or less. There is a big drop off in duration before six and a half and after 9. Perhaps if naps weren't included this histogram would look more like a normal distribution. Despite this, there still must be some error unaccounted for because I don't belive people are sleeping 16 hours straight.

```{r}
sleep_schedule |> 
  arrange(duration_hours)
```

I don't see a clear pattern by having a quick look.


```{r}
nap<-sleep_schedule |> filter(duration_hours<= 2)

# Using the sleep_schedule dataframe from the previous step
nap |> 
  ggplot(aes(x = bed_hour)) +
  # Set a single static color since we aren't comparing two events anymore
  geom_bar(fill = "#2c3e50", color = "black", alpha = 0.8) +
  # Force the x-axis to show every hour from 0 to 23
  scale_x_continuous(breaks = seq(0, 23, by = 1)) +
  labs(
    title = "Most Common Bedtimes",
    subtitle = "Based on minute-level sleep data (24-hour clock)",
    x = "Hour of Day (0 = Midnight, 12 = Noon)",
    y = "Total Number of Sleep Sessions"
  ) +
  theme_minimal()
```

I was hoping to see that the low durations of sleep corresponded times of day associated with nap time. It seems unlikely then that people are taking 2 hour naps in the middle of the night. It's more likely that the LogId that should be measuring the full sleep attempt is failing and loging more than one sleep attempt per night of sleep thus artificially lowering the durations of the main sleep of the day.


```{r}
sleep_schedule |> 
  # If you don't have a specific 'date' column, extract it from bedtime
  mutate(sleep_date = as.Date(bedtime)) |> 
  count(Id, sleep_date, name = "total_sessions") |> 
  # Sort to see the users with the most sessions per day at the top
  arrange(desc(total_sessions))
```

There aren't an inmense amount of breaks. 170 rows aprox out of 790 have more than one sleep_date. Maybe those can be the naps.

```{r}
sleep_schedule |> 
  filter(
    Id == 6117666160, 
    as.Date(bedtime) == "2016-03-16"
  ) |> 
  relocate(duration_hours, .before = bedtime) |> 
  relocate(Id, .after = day_of_week) 
```

Is this an error or does this guy have a terrible sleep schedule? In any case this tells me that having 2 LogIds in the same day doesn't necessaraly mean measurement error, nap or bad sleeping habits. They could also go to sleep after 0:00 in Tuesday (therefore is technically Wednesday) and before 0:00 the next day (also Wednesday).


```{r}
nap <- 
  sleep_schedule |> 
  filter(
    duration_hours <= 2, 
    between(hour(bedtime), 13, 15)
  )

not_nap <- sleep_schedule |> anti_join(nap)
  
```


```{r}
ggplot(not_nap, aes(x = duration_hours)) +

  # binwidth = 0.5 is crucial; it sets each bar to represent a 30-minute window.
  geom_histogram(
    binwidth = 0.5,
    fill = "#3498db",  
    color = "black",
    alpha = 0.8 
  ) +
  
  # Reference line (8-hour benchmark)
  geom_vline(
    xintercept = 8,
    linetype = "dashed",
    color = "#e74c3c", 
    linewidth = 1
  ) +
  
  # 4. Enhance readability and formatting
  scale_x_continuous(
    breaks = seq(0, max(sleep_duration$duration_hours, na.rm = TRUE) + 1, by = 1)
    ) +
  labs(
    title = "Sleep Duration Distribution",
    subtitle = "Histogram showing how often specific sleep lengths occur (in 30-minute bins)",
    x = "Total Sleep Duration (Hours)",
    y = "Frequency of sleep duration"
  ) +
  theme_minimal() +
  # Make the title and axes text clear for presentation
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 10, face = "italic"),
    axis.title = element_text(size = 11)
  )
```

Wasted my time on the nap thing. Very little impact on the distribution.

```{r timeHitSack}
sleep_schedule |>   
  # filter extreme cases simply using logic (NOT USING STATISTICS MAYBE CHANGE)
  filter(duration_hours<10) |> 
  
  # Extract just the hour to make it easy to find trends
  mutate(
    bed_hour = hour(bedtime),
    wake_hour = hour(wake_time)) |> 

  # Pivot the data to put both bed_hour and wake_hour into a single column
  pivot_longer(
    cols = c(bed_hour, wake_hour),
    names_to = "event_type",
    values_to = "hour"
  ) |> 
  
  # Rename the variables so the chart legend looks professional
  mutate(
    event_type = case_when(
      event_type == "bed_hour" ~ "Bedtime",
      event_type == "wake_hour" ~ "Wake Time"
    )
  ) |> 

  ggplot(aes(x = hour, fill = event_type)) +
  geom_bar(position = "dodge", color = "black", alpha = 0.8) +
  
  # Force the x-axis to show every hour from 0 to 23
  scale_x_continuous(breaks = seq(0, 23, by = 1)) +
  
  # Add custom Bellabeat brand colors
  scale_fill_manual(values = c("Bedtime" = "#2c3e50", "Wake Time" = "#e74c3c")) +
  
  labs(
    title = "Most Common Bedtimes and Wake Times",
    subtitle = "Each bar is the sum of all instances of an hour in the data. \n
    6:59 and 6:01 both count in the column of waking up at 6",
    x = "Hour of Day (0 = Midnight, 12 = Noon)",
    y = "Total Number of Sleep Sessions",
    fill = "Event"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```
This proves that naps are not common, yet late bedtimes are.

```{r bxpltDurationWkDay}
# Build the boxplot
ggplot(sleep_schedule, aes(x = day_of_week, y = duration_hours, fill = day_of_week)) +
  geom_boxplot(alpha = 0.7, color = "black") +
  
  # Add a visual benchmark for the recommended 8 hours
  geom_hline(yintercept = 8, linetype = "dashed", color = "#e74c3c", linewidth = 1) +
  
  labs(
    title = "Distribution of Sleep Duration by Day of the Week",
    subtitle = "Red dashed line represents the recommended 8 hours of sleep",
    x = "Day of the Week",
    y = "Total Sleep Duration (Hours)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") # Hides the legend since the x-axis already labels days
```

Doesn't really tell me much. Sure Sábado has a wider distribution with people willing to sleep less but not that much more than a random miércoles.

# 5. Visualizaciones de respaldo y hallazgos clave

## Sueño visualisations





# 6. Las tres recomendaciones de contenido de alto nivel basadas en tu análisis

# Dudas / apuntes

- Puede que tenga que utilizar logId
- cuando la gente se va a dormir, se despierta y si se echa la siesta parece interesante

Qué pregunta intentas responder y para quién. Cómo has limpiado y comprobado los datos. Qué conclusiones respaldan los resultados y qué limitaciones tienen. Qué recomendarías hacer y qué información adicional necesitaría
